What is the nature of decision noise in random exploration?

### 2022-11-21 draft/Older drafts/Older version/V3/Internal-External Noise.tex
Initial commit

\documentclass[12pt]{article}
\usepackage{times}
\usepackage{color}
\usepackage{apalike}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{amsmath}
%\usepackage[maxbibnames=99]{biblatex}
%\usepackage{setspace}
%\usepackage{geometry}
\usepackage[font={sf,small}]{caption}
%\usepackage{setspace}
%\usepackage{geometry}
%\usepackage{hyperref}
%\hypersetup{
%    colorlinks,
%    citecolor=black,
%    filecolor=black,
%    linkcolor=black,
%    urlcolor=black
%}
%\geometry{letterpaper}

@@

	\newpage
	\begin{abstract}

		The explore-exploit tradeoff is a fundamental behavioral dilemma faced by all adaptive organisms. Should we explore new options in the hopes of finding a better meal, a better house or a better mate, or should we exploit the options we currently believe to be best? Striking the right balance between exploration and exploitation is hard computational problem and there is significant interest in how humans and other animals make explore-exploit decisions in practice. One particularly effective strategy for solving the explore-exploit dilemma is choice randomization.  In this strategy, the decision process is noisy meaning that high value `exploit' options are not always chosen and exploratory choices are sometimes made by chance. In theory, such `random exploration', can be surprisingly effective in explore-exploit problems and, if implemented correctly, can come close to optimal performance. Recent work suggests that humans actually use random exploration to solve simple explore-exploit problems. Despite this progress a number of questions remain about the nature of random exploration as there are a number of ways in which seemingly stochastic choices could be generated. In one strategy, that we call the “external noise strategy”, participants could rely on stochasticity in the world and allow irrelevant features of the stimulus to drive choice. In another strategy called “internal noise strategy”, people could rely on stochastic processes within their own brains.  In this work, we modified our recently published `Horizon Task' in such a way as to distinguish these two strategies. Using both a model-free and model-based analysis of human behavior we show that, while both types of noise are present in explore-exploit decisions, random exploration is dominated by internal noise. This suggests that random exploration depends on adaptive noise processes in the brain which are subject to (perhaps unconscious) cognitive control. 
	\end{abstract}
	\newpage

@@

	Imagine  trying to decide where to go to dinner with a friend ... EXPLORE-EXPLOIT EXAMPLE.  More generally, such decisions are known as explore-exploit decisions ... there's considerable interest in how humans and animals solve it CITE 10 PAPERS INCLUDING WILSON ET AL. 2014...

	%The explore-exploit tradeoff is a fundamental behavioral dilemma faced by all adaptive organisms. Should we explore new options in the hopes of finding a better meal, a better house or a better mate, or should we exploit the options we currently believe to be best? Striking the right balance between exploration and exploitation is hard computational problem and there is significant interest in how humans and other animals make explore-exploit decisions in practice.

	One particularly effective strategy for solving the explore-exploit dilemma is choice randomization \cite{}. In this strategy, the decision process between exploration and exploitation is corrupted by `decision noise', meaning that high value `exploit' options are not always chosen and exploratory choices are sometimes made by chance. IN OUR RESTAURANT EXAMPLE ... In theory, such “random exploration”, is surprisingly effective and, if implemented correctly, can come close to optimal performance \cite{}.

	Recently we have shown that humans appear to actually use random exploration and actively adapt their decision noise to solve simple explore-exploit problems \cite{}.  The key manipulation in the task is the horizon condition, i.e. the number of decisions remaining to make. THE IDEA BEHIND THIS MANIPULATION IS THAT PEOPLE SHOULD EXPLORE MORE IN THE LONG HORIZON CONDITION (LINK BACK TO EXAMPLE).  USING SUCH A HORIZON MANIPULATION WE FOUND THAT  people have greater decision noise in the long versus the short horizon condition.

	However, a key limitation of this work was that the source of the decision noise used for exploration is unknown. Of particular interest is whether the adaptive decision noise that is linked to exploration is generated internally, within the brain, or arises externally, in the input from the world. Previous work makes a strong case for both types of noise being relevant to behavior. For instance, external, stimulus-driven noise is thought to be a much greater source of choice variability in perceptual decisions than internal noise \cite{}. Conversely internal, neural noise is thought to drive exploratory singing behavior in song birds \cite{} and the generation and control of this internal noise has been linked to specific neural structures.

	In this paper, we investigate which source of noise, internal vs external, drives random exploration in humans in a simple explore-exploit task adapted from our previous work \cite{}. To distinguish between the two types of noise, we had people make the exact same explore-exploit decision twice. If decision noise is purely externally driven, then people choices should be identical both times, that is their choices should be consistent since the stimulus is the same both times. Meanwhile, if noise is internally driven, the extent to which their choices are consistent should be determined by the level of the internal noise. By analyzing behavior on this task in both a model-free and model-based manner, we were able to show that, while both types of noise are present in explore-exploit decisions, the contribution of internal noise to random exploration far exceeds that contributed by the stimulus.

@@

	\subsection*{The Repeated-Games Horizon Task}
	We used a modified version of our previously published `Horizon Task' (Figure \ref{fig:task}) to show the influence of internal vs external noise on people's decisions (CITE). THE KEY MANIPULATION WAS REPEATED GAMES SAME DECISION TWICE - LINK INTO EXAMPLE

	MORE SPECIFICALLY, 
	In this task, participants play a set of games in which they make choices between two slot machines (one-armed bandits) that pay out rewards from different Gaussian distributions. To maximize their rewards in each game, participants need to exploit the slot machine with the highest mean, but they cannot identify this best option without exploring both options first.

	%Participants played a series of games of a sequential two-armed bandit gambling task. 
	In each game they made multiple decisions between two options. Each option paid out a random reward between 0 and 100 points sampled from a Gaussian distribution. The means of the underlying Gaussian were different for the two bandit options, remained the same within a game, but changed with each new game. One of the bandit will always have a higher mean than the other. Participants were instructed to maximize the points earned over the entire task.

	The first four trials of each game were forced-choice trials, in which only one of the options was available for the participant to choose. We used these forced-choice trials to manipulate the relative ambiguity of the two options, by providing the participant with different amounts of information about each bandit before their first free choice. The four forced-choice trials set up two uncertainty conditions: unequal uncertainty(or [1 3]) in which one option was forced to be played once and the other three times, and equal uncertainty(or [2 2]) in which each option was forced to be played twice. After the forced-choice trials, participants made either 1 or 6 free choices (two horizon conditions).

	THESE CONDITIONS ALLOW US TO MEASURED DIRECTED AND RNADOM EXPLORATION IN A MODEL-FREE WAY.  REPLICATION OF PREVIOUS RESULTS

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/taskfig.PNG}
			\caption{A key additional manipulation here is repeated games. Each pair of repeated games with identical example trials will appear twice during the experiment. We setup the repeated games such that they are at least 5 games apart from each other. A model-free measure of choice inconsistency which reflex the underlying decision noise is defined as the proportion of disagreement for repeated games.}
			\label{fig:task2}
		\end{center}
	\end{figure}

	Finally, the crucial additional manipulation in this task is repeated games (Figure \ref{fig:task2}). In each pair of repeated games, the four forced-choice trials were yoked, meaning that on the first free choice trial participants were faced with identical stimuli.  After the first free choice trial, the outcomes on the repeated games were not yoked and the outcomes were sampled independently from the underlying Gaussian distribution.  Not yoking the later trials made it harder for participants to detect repeated games.  In addition, the presentation of repeated games was controlled so that each repeated pair was at least five games away from each other.

@@

	\subsubsection*{Model-free analysis}

	In the model-free analysis we asked whether participants' choices were  consistent or inconsistent in the two repetitions of each game.  The idea behind this measure is that purely external noise should lead to consistent choices as the external stimulus is identical both time.  Conversely, internal noise should lead to inconsistent choices both times.  SPELL OUT MORE

	%used choice inconsistency in the repeated games as a measure of internal noise.

	%GOT TO INTRODUCE DIRECTED EXPLORATION - MAYBE BEFORE THIS

	%MOVE THIS PARAGRAPH UP As in our previous paper \cite{}, the influence of directed exploration is measured as the probability of choosing the more informative option, i.e. the bandit with only 1 outcome in [1 3] condition; decision noise is measured as the probability of choosing the low mean option in [2 2] condition.

	In addition, whether participants make consistent choices in repeated games is used as a model-free measure of internal noise, since in repeated trials, external noise should be identical on both trials. So only internal noise can differ and drive the choice inconsistency. The degree to which people make consistent choices in repeated trials can reflect the internal noise.

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/Modelfree.png}
			\caption{Both direct (P $<$ 0.001) and random exploration (P $<$ 0.001 for both [1 3] and [2 2]) increases with horizon.(A,B) Choice consistency in both [1 3] and [2 2] condition increases with horizon(P([1 3]) $<$ 0.001, P([2 2])$<$0.001).(C)}
			\label{fig:mf}
		\end{center}
	\end{figure}

	%In line with previous findings, both directed and random exploration increase with horizon (Figure \ref{fig:mf}). In addition, 
	Choice inconsistency between repeated games was non-zero in both horizon conditions, suggesting that not all of the noise was stimulus driven.  In addition, choice inconsistency was higher in horizon 6 than in horizon 1 for both [1 3] and [2 2] condition (Figure \ref{fig:mf}), suggesting that at least some of the horizon dependent noise is internal.

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/modelfree_intext.png}
			\caption{For both [1 3] and [2 2] condition, there is a significant difference between people's behavior and predicted choice inconsistency assuming that only external noise exists where people should behave identically in repeated games. Also, there is a significant difference between people's behavior and predicted choice inconsistency assuming that only internal noise exists where people treat repeated games independently. }
			\label{fig:mf2}
		\end{center}
	\end{figure}

	To gain more quantitative insight into the data we computed predicted values of the choice inconsistency for the purely external and purely internal noise cases.  For the purely externally driven noise case, then people should make the exact same decisions each time in repeated games, which corresponds to the light blue line in Figure \ref{fig:mf2}. If noise is however purely internally driven, then there should be no stimulus-dependent noise component such that people should in principle treat repeated games independently, hence 
	\begin{equation*}
	\begin{split}
	P(\mbox{consistency}) &= P(\mbox{low mean})^2 + P(\mbox{high mean})^2\\
	&= P(\mbox{low mean})^2 + (1-P(\mbox{low mean}))^2\\ \mbox{hence},\quad P(\mbox{inconsistency}) &= 1 - P(\mbox{consistency})
	\end{split}
	\end{equation*},

	we can make the prediction of p(disagree) if there is only internal noise. However, people's behavior falls in between the pure external noise prediction and the pure internal noise prediction, suggesting that both external and internal noise are present in driving this choice inconsistency.

	Since choice inconsistency only reflects internal noise, Figure \ref{fig:mf2} suggests that internal noise increases with horizon, however we can not draw any obvious conclusions about whether external noise is horizon dependent or not.

	\subsubsection*{Model-based analysis}
	To more precisely quantify the size of internal and external noise in this task, we turned to model fitting.

	\paragraph{Overview of model}
	As with our model-free analysis, the model-based analysis focuses only on the first free-choice trial since that is the only free choice when we have control over the information bias between the two bandits. To model participants’ choices on this first free-choice trial, we assume that they make decisions by computing the difference of value $\Delta Q$ between the two options:

	$$\Delta Q= \Delta R+A \Delta I+b+n_{ext}+n_{int}$$

	Where, the experimentally controlled variables are $\Delta R=R_{right}-R_{left}$, difference between the mean of rewards shown on the forced trials, and $\Delta I$, the difference information available for playing the two options on the first free-choice trial (for simplicity we define $\Delta I$ to be +1, -1 or 0: +1 if one reward is drawn from the right option and three are drawn from the left in the [1 3] condition, -1 if one from the left and three from the right, and in [2 2] condition, $\Delta I$ is 0). $n_{ext}$ and $n_{int}$ are external noise and internal noise respectively.

	The subject-and-condition-specific parameters are: the spatial bias, b, which determines the extent to which participants prefer the option on the right, the information bonus A, which controls the level of directed exploration, ϵ denotes the external, external noise, which is identical on the repeat versions of each game, and n denotes internal noise, which is uncorrelated between repeat plays. For repeated games, the set of forced-choice trials are exact the same, so the external noise should be the same, but the free decision noise may be different. This is exactly how we distinguish external noise from internal noise. In symbolic terms, for repeated games i and j,  $\sigma_{ext}^i=\sigma_{ext}^j$  and $\sigma_{int}^i \neq \sigma_{int}^j$.
	The decision model assumes that participants choose right if $\Delta Q > 0$, otherwise the participant will choose the left option.

	We used hierarchical Bayesian analysis to fit the parameters (see Figure \ref{fig:model} for an graphical representation of the model in the style of [5]). In particular, we fit values of the information bonus A, spatial bias B, variance of internal noise $\sigma_{int}^2$, and variance of external noise, $\sigma_{ext}^2$ for each participant in each horizon and uncertainty condition. The mean and standard deviation of information bonus A and spatial bias B are sampled from a Gaussian prior and an exponential prior respectively. The variance for both type of noises were sampled from a gamma distribution, and the group-level parameter k and λ for the gamma distribution are sampled from exponential priors. At the game level, we sampled one stimulus driven noise for each pair of repeated games and we sampled two free decision noises for each individual game in the pair.  The model in Figure \ref{fig:model} is fitted using the MATJAGS and JAGS \cite{}.

	\begin{table}[H]
		\small
		\begin{center}
			\begin{tabular}{|c|c|c|c|}
				\hline
				Parameter &
				Horizon dependent? &
				Uncertainty dependent? &
				Game dependent? \\
				\hline
				information bonus, $A$ &
				yes &
				n/a &
				no \\
				\hline
				spatial bias, $B$ &
				yes &
				yes &
				no \\
				\hline
				external decision noise, $\sigma_{ext}$ &
				yes & 
				yes &
				no \\
				\hline
				internal decision noise, $\sigma_{int}$ &
				yes & 
				yes &
				yes \\
				\hline
			\end{tabular}
		\end{center}
		\caption{Model parameters. }
		\label{table1}
	\end{table}%

@@

	\paragraph{Model fitting results} 
	%Posterior distributions over the group-level means are shown in the left column of Figure \ref{fig:mb1}, while posteriors over the TMS-related change in parameters are shown in the right column.  Both columns suggest a selective effect of RFPC stimulation on the information bonus in horizon 6.

	Only internal noise increases  with horizon (Figure \ref{fig:modelbased}). 
	%In both horizon 1 and horizon 6, internal noises are significantly higher than internal noises (for both horizon 1 and horizon 6, P $<$ 0.001).

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=0.5\textwidth]{figures/bayes.png}
			\caption{The change of external and internal decision noises between horizons. The errorbars are confidence intervals at $95\%$. Only internal noise increases significantly with horizon.}
			\label{fig:modelbased}
		\end{center}
	\end{figure}

	Figure 3 also demonstrates the existence of both external noise and internal noise. Hierarchical Bayesian analysis provides us a way to quantify the magnitude of external and internal noise respectively, from the result, internal noise is the main source of random exploration.

	\section*{Discussion}
	In this paper, we investigated whether random exploration is driven by internal noise, putatively arising in the brain, or external noise, arising from the environment.  We find horizon dependent changes in both internal and external sources of noise, but that the effect of internal noise is much greater. 
	One limitation of this work is in the interpretation of the different noises as being internal and external. In particular, while we controlled many aspects of the stimulus across repeated games (e.g. the outcomes and the order of the forced trials), we could not perfectly control all stimuli the participant received which, for example, would vary based on where they were looking.  Thus, our estimate of external noise is likely a lower bound. Likewise, our estimate of internal noise is likely an upper bound as these “missing” sources of stimulus driven noise would be interpreted as internal noise in our model. Despite this, the sheer magnitude of the difference between internal and external noise (internal noise is 2-3 times the size of external noise, Figure 3), suggests that our interpretation may be safe as an awful lot of external noise would have to be explained by variables not relevant to the task.  
	The horizon-dependent increase in internal noise is consistent with the idea that random exploration is driven by intrinsic variability in the brain. This is in line with work in the bird song literature in which song variability during song learning has been tied to neural variability arising from specific areas of the brain. In addition, this work is consistent with a recent report from Ebitz et al [7] in which behavioral variability in an “explore” state was also tied to internal rather than external sources of noise. Whether such a noise-controlling area exists in the human brain is less well established, but one candidate theory [4] suggests that norepinephrine (NE) from the locus coeruleus may play a role in modulating internal levels of noise.  While there is some evidence that NE plays a role in explore-exploit behavior [8], this link has been questioned [9].

>>>>

\documentclass[12pt]{article}
\usepackage{hyperref}
\usepackage[authoryear, round,sort,comma,numbers]{natbib}
\usepackage{times}
\usepackage{color}
\usepackage{apalike}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{amsmath}

%\usepackage[maxbibnames=99]{biblatex}
%\usepackage{setspace}
%\usepackage{geometry}
\usepackage[font={sf,small}]{caption}
%\usepackage{setspace}
%\usepackage{geometry}
%\usepackage{hyperref}
%\hypersetup{
%    colorlinks,
%    citecolor=black,
%    filecolor=black,
%    linkcolor=black,
%    urlcolor=black
%}
%\geometry{letterpaper}

@@

	\newpage
	\begin{abstract}
		The explore-exploit tradeoff is a fundamental behavioral dilemma faced by all adaptive organisms, from everyday life decisions like deciding for a meal to important life decisions like finding a life partner. It is computationally a hard problem to find the right balance between exploration and exploitation and hence there is significant interest in how humans and other animals solve the explore-exploit dilemma. One particularly effective strategy for solving the explore-exploit dilemma is choice randomization.  In this strategy, the decision process is noisy meaning that high value `exploit' options are not always chosen and exploratory choices are sometimes made by chance. In theory, such `random exploration', can be surprisingly effective in explore-exploit problems and, if implemented correctly, can come close to optimal performance. Recent work suggests that humans actually use random exploration to solve simple explore-exploit problems. Despite this progress a number of questions remain about the nature of random exploration as there are a number of ways in which seemingly stochastic choices could be generated. In one strategy, that we call the “external noise strategy”, participants could rely on stochasticity in the world and allow irrelevant features of the stimulus to drive choice. In another strategy called “internal noise strategy”, people could rely on stochastic processes within their own brains. In this work, we modified our recently published `Horizon Task' in such a way as to distinguish these two strategies. Using both a model-free and model-based analysis of human behavior we show that, while both types of noise are present in explore-exploit decisions, random exploration is dominated by internal noise. This suggests that random exploration depends on adaptive noise processes in the brain which are subject to (perhaps unconscious) cognitive control. 
	\end{abstract}
	\newpage

@@

	Imagine trying to decide where to go to dinner with a friend, you can go to your favorite restaurant that you both really enjoy and always go to, or you can try the new restaurant that just opened a few days ago right across the street to the other restaurant which may end up being your newly favorite. More generally, such decisions are known as explore-exploit decisions. The explore-exploit decision refers to deciding between exploiting the best known option so far, like going to your old favorite restaurant, and exploring other options for potential better decisions in the future, like trying the new restaurant. There's considerable interest in how humans and animals solve it.\citep{eegittins74, eeauer02,eegittins79,eekrebs78,eethompson33, eewatkins89, eebridle90,eemeyer95, eebanks97,eefrank09, eesteyvers09, eelee11, eepl12,eezhang13,eedaw06, eepl11, wilson2014}

	One particularly effective strategy for solving the explore-exploit dilemma is choice randomization \citep{eethompson33, eewatkins89, eebridle90}. In this strategy, the decision process between exploration and exploitation is corrupted by `decision noise', meaning that high value `exploit' options are not always chosen and exploratory choices are sometimes made by chance. In our restaurant example, your restaurant decision does not always depend on the quality of the food, you are very likely to go to the new restaurant if you happen to see another old friend walking right in, or you wait until the last moment to make a split second decision about where to go as if you flipped a mental coin in your head and decide to go to your old favorite if it's heads up. In theory, such “random exploration”, is surprisingly effective and, if implemented correctly, can come close to optimal performance \citep{eebridle90}.

	Recently we have shown that humans appear to actually use random exploration and actively adapt their decision noise to solve simple explore-exploit problems \citep{wilson2014}.  The key manipulation in the task is the horizon condition, i.e. the number of decisions remaining to make. The idea behind this manipulation is that people should explore more in the long horizon condition. If you are leaving town tomorrow for vacation, you probably want to go to your old favorite to guarantee a good last meal, but if you are not going anywhere, you would be more likely to try the new restaurant. Using such a horizon manipulation we found that people have greater decision noise in the long versus the short horizon condition.

	However, a key limitation of this work was that the source of the decision noise used for exploration is unknown. Of particular interest is whether the adaptive decision noise that is linked to exploration is generated internally, within the brain, or arises externally, in the input from the world. In the restaurant example, an old friend walking by would be a source of external noise, but flipping a mental coin would be an internal noise. Previous work makes a strong case for both types of noise being relevant to behavior. For instance, external, stimulus-driven noise is thought to be a much greater source of choice variability in perceptual decisions than internal noise \citep{eeBrunton13}. Conversely internal, neural noise is thought to drive exploratory singing behavior in song birds  \citep{songbird2} and the generation and control of this internal noise has been linked to specific neural structures.

	In this paper, we investigate which source of noise, internal vs external, drives random exploration in humans in a simple explore-exploit task adapted from our previous work \citep{wilson2014}. To distinguish between the two types of noise, we had people make the exact same explore-exploit decision twice. If decision noise is purely externally driven, then people choices should be identical both times, that is their choices should be consistent since the stimulus is the same both times. Meanwhile, if noise is internally driven, the extent to which their choices are consistent should be determined by the level of the internal noise. By analyzing behavior on this task in both a model-free and model-based manner, we were able to show that, while both types of noise are present in explore-exploit decisions, the contribution of internal noise to random exploration far exceeds that contributed by the stimulus.

	%\begin{figure}[H]
	%	\begin{center}
	%		\includegraphics[width=0.7\textwidth]{figs/bayes_s.png}
	%			\caption[Subject level estimated mean of external and internal noise variance.]{Subject level estimated mean of external and internal noise variance. Both internal and external noise increase with horizon, internal noise is dominant.}
	%			\label{fig:mb3}
	%		\end{center}
	%	\end{figure}

@@

	\subsection*{The Repeated-Games Horizon Task}
	We used a modified version of our previously published `Horizon Task' \citep{wilson2014} to show the influence of internal vs external noise on people's decisions (Figure \ref{fig:taskfig}). The key manipulation was to use repeated games to let people make the same decision twice. In the restaurant example, if your decision is mainly driven by external noise, then a few months later when you see the friend walking in front of you into the restaurant, you are very likely to make the same decision and follow him into that restaurant and says hi. However, if your decision is mainly driven by internal noise, then next time you make a split-second decision about where to go, you are equally likely to go to the other restaurant.

	More specifically, we look at the contribution of external and internal noise by providing with participants the same decision problem twice in the task. In this task, participants play a set of games in which they make choices between two slot machines (one-armed bandits) that pay out rewards from different Gaussian distributions. To maximize their rewards in each game, participants need to exploit the slot machine with the highest mean, but they cannot identify this best option without exploring both options first.

	In each game they made multiple decisions between two options. Each option paid out a random reward between 1 and 100 points sampled from a Gaussian distribution. The means of the underlying Gaussian were different for the two bandit options, remained the same within a game, but changed with each new game. One of the bandit will always have a higher mean than the other. Participants were instructed to maximize the points earned over the entire task.

	The first four trials of each game were forced-choice trials, in which only one of the options was available for the participant to choose. We used these forced-choice trials to manipulate the relative ambiguity of the two options, by providing the participant with different amounts of information about each bandit before their first free choice. The four forced-choice trials set up two uncertainty conditions: unequal uncertainty(or [1 3]) in which one option was forced to be played once and the other three times, and equal uncertainty(or [2 2]) in which each option was forced to be played twice. After the forced-choice trials, participants made either 1 or 6 free choices (two horizon conditions). In unequal uncertainty condition, people are more likely to choose the option that they know less about - the more informative option - to explicitly explore that option more. This type of information driven exploration is known as directed exploration.

	As in our previous paper \citep{wilson2014}, These conditions allow us to measure directed and random exploration in a model-free way. Directed exploration is measured as the probability of choosing the more informative option in [1 3] condition whereas random exploration is measured as the probability of choosing the low mean option in [2 2] condition. In line with previous results \citep{wilson2014}, we showed that both directed and random exploration increase with horizon. Both direct (t(59)=6.88, p $<$ 0.001) and random exploration (t(59) = 6.17, p $<$ 0.001 for [1 3], t(59) = 7.26, p $<$ 0.001 for [2 2]) increases with horizon. (Figure \ref{fig:modelfree}, A,B).   Directed exploration is considered to be driven by information bias and random exploration is considered to be driven by decision noise, in this work, we are investigating which source of decision noise, external vs internal, drives random exploration.

	%the influence of directed exploration is measured as the probability of choosing the more informative option, i.e. the bandit with only 1 outcome in [1 3] condition; decision noise is measured as the probability of choosing the low mean option in [2 2] condition.

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/taskfig.PNG}
			\caption{A key additional manipulation here is repeated games. Each pair of repeated games with identical example trials will appear twice during the experiment. We setup the repeated games such that they are at least 5 games apart from each other. A model-free measure of choice inconsistency which reflex the underlying decision noise is defined as the proportion of inconsistent choices for repeated games.}
			\label{fig:taskfig}
		\end{center}
	\end{figure}

	Finally, the crucial additional manipulation in this task is repeated games (Figure \ref{fig:taskfig}). In each pair of repeated games, the four forced-choice trials were yoked, meaning that on the first free choice trial participants were faced with identical stimuli.  After the first free choice trial, the outcomes on the repeated games were not yoked and the outcomes were sampled independently from the underlying Gaussian distribution.  Not yoking the later trials made it harder for participants to detect repeated games.  In addition, the presentation of repeated games was controlled so that each repeated pair was at least five games away from each other.

@@

	\subsubsection*{Model-free analysis}
	In the model-free analysis we asked whether participants' choices were  consistent or inconsistent in the two repetitions of each game. The idea behind this measure is that purely external noise should lead to consistent choices as the external stimulus is identical both time. Conversely, internal noise should lead to independent choices, and hence possible inconsistent choices both times. More specifically, we look at the proportion of times that a participant make inconsistent decisions in repeated game. Choice inconsistency is defined as the proportions of inconsistent choices for repeated games. (See Figure \ref{fig:taskfig})

	In addition, whether participants make consistent choices in repeated games is used as a model-free measure of internal noise, since in repeated trials, external noise should be identical on both trials. So only internal noise can differ and drive the choice inconsistency. The degree to which people make consistent choices in repeated trials can reflect the internal noise.  Since choice inconsistency in both [1 3](t(59) = 5.10, P([1 3]) $<$ 0.001) and [2 2](t(59) = 6.08, P([2 2])$<$0.001) condition increase with horizon (Figure \ref{fig:modelfree} C), this is a behavioral evidence that internal noise increases with horizon.

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/modelfree.png}
			\caption{Both directed and random exploration increase with horizon. Choice inconsistency also increases with horizon for both [1 3] and [2 2] conditions. }
			\label{fig:modelfree}
		\end{center}
	\end{figure}

	Choice inconsistency between repeated games was non-zero in both horizon conditions, suggesting that not all of the noise was stimulus driven.  In addition, choice inconsistency was higher in horizon 6 than in horizon 1 for both [1 3] and [2 2] condition (Figure \ref{fig:modelfree}), suggesting that at least some of the horizon dependent noise is internal.

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/modelfree_intext.png}
			\caption[Both external and internal noise contribute to the choice variability in random exploration]{Both external and internal noise contribute to the choice variability in random exploration. For both [1 3] and [2 2] condition, there is a significant difference between people's behavior and predicted choice inconsistency assuming that only external noise exists where people should behave identically in repeated games. Also, there is a significant difference between people's behavior and predicted choice inconsistency assuming that only internal noise exists where people treat repeated games independently. }
			\label{fig:mf2}
		\end{center}
	\end{figure}

	To gain more quantitative insight into the data we computed predicted values of the choice inconsistency for the purely external and purely internal noise cases.  For the purely externally driven noise case, then people should make the exact same decisions each time in repeated games, so $P(\mbox{inconsistent})$ should be 0, which corresponds to the light blue line in Figure \ref{fig:mf2} in [1 3] and [2 2] respectively. If noise is however purely internally driven, then there should be no stimulus-dependent noise component such that people should in principle treat repeated games independently, and the extent to which they are consistent with themselves in repeated games can be predicted by choosing the low mean option twice or choosing the high mean option twice, which corresponds to the red line in Figure \ref{fig:mf2}.

	\begin{equation*}
	\begin{split}
	P(\mbox{consistent}) &= P(\mbox{low mean})^2 + P(\mbox{high mean})^2\\
	&= P(\mbox{low mean})^2 + (1-P(\mbox{low mean}))^2\\ \mbox{hence},\quad P(\mbox{inconsistent}) &= 1 - P(\mbox{consistent})
	\end{split}
	\end{equation*}

	However, people's behavior falls in between the pure external noise prediction and the pure internal noise prediction (Figure \ref{fig:mf2}), suggesting that both external and internal noise are present in driving this choice inconsistency. Since choice inconsistency only reflects internal noise, Figure \ref{fig:mf2} suggests that internal noise increases with horizon.

	\subsubsection*{Model-based analysis}
	To more precisely quantify the size of internal and external noise in this task, we turned to model fitting. We modeled behavior on the first free choice of the Horizon Task using a version of the logistic choice model in \citep{wilson2014} that was modified to differentiate internal and external noises. In particular, we assume that in repeated games, external noise remains the same whereas internal noise can change.

	\paragraph{Overview of model}
	As with our model-free analysis, the model-based analysis focuses only on the first free-choice trial since that is the only free choice when we have control over the information bias between the two bandits. To model participants’ choices on this first free-choice trial, we assume that they make decisions by computing the difference of value $\Delta Q$ between the right and left options, choosing right when $\Delta Q > 0$ and left otherwise.  Specifically, we write

	\begin{equation}
	\Delta Q= \Delta R+A \Delta    I+b+n_{ext}+n_{int}
	\end{equation}

	Where, the experimentally controlled variables are $\Delta R=R_{right}-R_{left}$, the difference between the mean of rewards shown on the forced trials, and $\Delta I$, the difference information available for playing the two options on the first free-choice trial. For simplicity, and because information is manipulated categorically in the Horizon Task, we define $\Delta I$ to be +1, -1 or 0, +1 if one reward is drawn from the right option and three are drawn from the left in the [1 3] condition, -1 if one from the left and three from the right, and in [2 2] condition, $\Delta I$ is 0. $n_{ext}$ and $n_{int}$ are external noise and internal noise respectively.

	The subject-and-condition-specific parameters are: the spatial bias, $b$, which determines the extent to which participants prefer the option on the right; the information bonus $A$, which controls the level of directed exploration; $n_{ext}$ denotes the external, external noise, which is identical on the repeat versions of each game; and $n_{int}$ denotes internal noise, which is uncorrelated between repeat plays and changes every game.

	For each pair of repeated games, the set of forced-choice trials are exactly the same, so the external noise, $n_{ext}$, should be the same while the internal noise, $n_{int}$ may be different. This is exactly how we distinguish external noise from internal noise. In symbolic terms, for repeated games i and j,  $n_{ext}^i=n_{ext}^j$  and $n_{int}^i \neq n_{int}^j$.

	\paragraph{Model fitting}
	We used hierarchical Bayesian analysis to fit the parameters of the model (see Figure \ref{fig:model} for an graphical representation of the model in the style of \cite{lee2014}). In particular, we fit values of the information bonus $A$, spatial bias $B$, variance of internal noise $\sigma_{int}^2$, and variance of external noise, $\sigma_{ext}^2$ for each participant in each horizon. The mean and standard deviation of information bonus $A$ and spatial bias $B$ are sampled from a Gaussian prior and an exponential prior respectively. The variance for both type of noises were sampled from a gamma distribution, and the group-level parameter $k$ and $\lambda$ for the gamma distribution are sampled from exponential priors.

	To capture the idea that external noise should be identical on repeated games, we sampled one value of the external noise, $n_{ext}$ for each pair of repeated games.  Conversely, because internal noise is expected to change between games we sampled two values of the internal noise, $n_{int}^1$ and $n_{int}^2$, i.e. one for each individual game.

	The model in Figure \ref{fig:model} is fitted using the MATJAGS and JAGS software \citep{jags, matjags}.

@@

	\paragraph{Model fitting results}

	%Group-level estimates for the variances of internal and external noise are shown in Figure \ref{fig:modelbased}A.  While both variances are non-zero, this shows that the internal noise is much larger than external noise. This horizon-based change is probed further in Figure XXXB in which we plot the posterior distributions over the change in internal and external noise with horizon.  This clearly shows that only internal noise varies with horizon (STATS XXX\% of the samples above zero for internal noise, XXX\% of samples above zero for external noise).

	Posterior distributions over the group-level means of the external and internal noise variance are shown in Figure \ref{fig:mb1}. While both variances are non-zero, this shows that the internal noise is much larger than external noise. This horizon-based change is probed further in Figure \ref{fig:mb2} in which we plot the posterior distributions over the change in internal and external noise with horizon. As is clear from this plot, internal noise increases dramatically with horizon whereas there is a slight increase of external noise with horizon as well (0 just falls in the critical region at level of 5\% for external noise but lies right at the border, although significant, it's not as strong as the internal noise). This clearly shows that internal noise dominates and significantly varies with horizon  (100\% of the samples above zero for internal noise, 96.33\% of samples above zero for external noise).

	%Furthermore, we plotted the subject level estimates of both internal and external noises in a similar fashion of the model-free plot of choice-inconsistency for comparison (Figure \ref{fig:mb3}). We showed again that both internal and external noises increase significantly with horizon. But in terms of magnitude, internal noise is what' really dominant here (See also Figure \ref{fig:rcurvepd6} in Appendix \ref{ch:appendicies:reward}). Subject-level estimates also allow us to look at the relationship between internal and external noises, and we showed that internal noise correlates with external noise significantly in both horizon conditions ($R^2 = 0.32, p < 0.001$ in horizon 1, $R^2 = 0.24, p < 0.001$ in horizon 6, see Figure \ref{fig:corrintext} in Appendix \ref{ch:appendix:corrintext}).

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=0.7\textwidth]{figures/dist_hypernoiseh.png}
			\caption[Posterior distributions over the group-level means of the external and internal noise variance.]{Posterior distributions over the group-level means of the external and internal noise variance. Both internal and external noises are nonzero, and internal noise has a much greater magnitude.}
			\label{fig:mb1}
		\end{center}
	\end{figure}

	%LEFT EACH TYPE OF NOISE IN EACH HORIZON
	%RIGHT IS DISTRIBUTION OVER CHANGE IN NOISE WITH HORIZON

	%Posterior distributions over the group-level means of the external and internal noise variance are shown in Figure \ref{fig:modelbased}.  As is clear from this plot, only internal noise increases  with horizon. 
	%In both horizon 1 and horizon 6, internal noises are significantly higher than internal noises (for both horizon 1 and horizon 6, P $<$ 0.001).

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=0.7\textwidth]{figures/dist_hypernoise.png}
			\caption[Posterior distributions over the group-level means of the change of external and internal noise variance.]{Posterior distributions over the group-level means of the change of external and internal noise variance. Internal noise increases significantly with horizon, external noise increases with horizon as well.}			
			\label{fig:mb2}
		\end{center}
	\end{figure}
